### 2010-10-19 Thesis/Chapter2.tex
- Update 12 October -
memoir class used now, fancy chapter style



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@185 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
- Update 19 October -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@186 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Background in continuum mechanics for soft-tissue modelling}
\label{chap2}
\begin{shortAbstract}
As seen in the previous chapter, realistic modelling of organ deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedures and per-operative guidance based on simulation. In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations. If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial. Therefore, a substantial comprehension of the mechanics involved and a knowledge of the physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs. This chapter will start by introducing the main concepts of continuum mechanics that are fundamental to study the mechanical response of organs. It will then present mathematical models able to describe the different mechanical aspects of materials and briefly discuss the mechanical characterisation of tissues.
\end{shortAbstract}

@@

\section{Description of motion} 	\label{chap2:descriptionMotion}
Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$. For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}. The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}. A change in the configuration of a continuum body results in a displacement. The displacement of a body has two components: a rigid-body displacement and a deformation. A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size. \ON In constrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation. This new configuration $\kappa$ may be designated by the \emph{current} or \emph{deformed configuration}. \OFF

@@

	\subsection{Lagrangian description}	
In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$. \ON Consequently, the current coordinates ($\mathbf{x} \in \kappa$) depend on the reference coordinates ($\mathbf{X} \in \kappa_{0}$): \OFF
\begin{equation}
\textbf{x} = \chi(\textbf{X}, t), \quad \chi(\textbf{X}, 0) = \textbf{X},
\end{equation}
\ON and a typical variable $\phi$ defined over the body is expressed in terms of the coordinates $\mathbf{X}$ and time $t$: \OFF
\begin{equation}
\label{chap2:phi}
\phi = \phi(\textbf{X}, t).
\end{equation}
\ON Let us consider a particule $X$ whose position in the reference configuration is $\mathbf{X}$. From \eqref{chap2:phi}, we note that the value of $\phi$ associated with this fixed particle $X$ changes over time. Hence, \OFF the Lagrangian description focuses its attention on the particles of the continuous body and it is usually used in solid mechanics.

	\subsection{Eulerian description}
\ON In constrast, the Eulerian description is interested in changes at fixed locations. This time, the motion and $\phi$ are described with respect to the current position ($\mathbf{x} \in \kappa$): \OFF
\begin{equation}
\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t).
\end{equation}
For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with the particle occupying the position $\textbf{x} \in \kappa$, which may very well be a different particle for each new time $ t $. Because a change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particle, the Eulerian description is focused on a spatial position. This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}. Because we are only interested in the study of solid bodies, the Lagrangian description will be used in the rest of the text.

	\subsection{Displacement field}
The displacement $ \mathbf{u} $ of a particle $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is given by the following:
\begin{equation}
\mathbf{u}(\textbf{X}, t) = \mathbf{x}(\mathbf{X}, t) - \mathbf{X}.
\end{equation}
A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body. \ON Thus, the deformed configuration $\kappa$ may be obtained from the undeformed configuration $\kappa_{0}$ by merely adding the displacement field: \OFF $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.

@@

	\subsection{Deformation gradient tensor}
The displacement field tells us how a \ON particule \OFF displaces from the reference to the deformed configuration. \ON More interestingly, we would like to know how a material line would deform (stretch and rotation) within this displacement field. \OFF Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$. \ON We now seek the relation existing between $d\mathbf{x}$ in the deformed configuration and $d\mathbf{X}$ of the reference configuration. \OFF

@@

	\subsection{Change of volume}
At this point, we have seen how a deformation can affect a vector. We will now look into its effect on volumes. Our motivation comes from the need to write global equilibrium statements that involve integrals over volumes. \ON Let us consider a volume in the reference configuration formed by \OFF three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ as can be seen on \fig{chap2:fig-changeVolume}.
% 
\begin{figure}[ht]
\begin{center}
\includegraphics[width=12cm]{chapter2/changeVolume.pdf}
\end{center}
\caption{Transformation of a volume element under a deformation mapping.}
\label{chap2:fig-changeVolume}
\end{figure}
%

The three vectors after deformation $d\mathbf{x}^{(1)}$, $d\mathbf{x}^{(2)}$ and $d\mathbf{x}^{(3)}$ can be obtained with:
\begin{equation}
\label{chap2:dxdX}
d\mathbf{x}^{(i)} = \mathbf{F} \, d\mathbf{X}^{(i)}, \quad i = 1, 2, 3.
\end{equation}
The volume of the parallelepiped that we will note $dV$  can be calculated using the triple product between the three vectors:
\begin{align}
dV &= d\mathbf{X}^{(1)} \cdot d\mathbf{X}^{(2)} \times d\mathbf{X}^{(3)} = (\mathbf{\hat{N}}_{1} \cdot \mathbf{\hat{N}}_{2} \times \mathbf{\hat{N}}_{3}) \, dX^{(1)} dX^{(2)} dX^{(3)} \notag \\
&= dX^{(1)} dX^{(2)} dX^{(3)}, \label{chap2:dV}
\end{align}
where $\mathbf{\hat{N}}_{i}$ \ON is \OFF the unit vector along $d\mathbf{X}^{i}$. \ON Similarly, we can compute the \OFF corresponding volume in the deformed configuration by:
\begin{align}
dv &= d\mathbf{x}^{(1)} \cdot d\mathbf{x}^{(2)} \times d\mathbf{x}^{(3)} \notag \\
&=  (\mathbf{F} \cdot \mathbf{\hat{N}}_{1}) \cdot (\mathbf{F} \cdot \mathbf{\hat{N}}_{2}) \times (\mathbf{F} \cdot \mathbf{\hat{N}}_{3}) \,  dX^{(1)} dX^{(2)} dX^{(3)} \quad \text{by \eqref{chap2:dxdX}} \notag \\
&= \det \mathbf{F} \, dX^{(1)} dX^{(2)} dX^{(3)}. \label{chap2:dv}
\end{align}
The determinant of $\mathbf{F}$ is called the \emph{Jacobian} and it is denoted by $J = \det \mathbf{F}$. \ON Therefore, the relation between the deformed volume $ dv $ and the undeformed volume $ dV $ may be written as: \OFF
\begin{equation}
\label{chap2:dvdV}
dv = J dV.
\end{equation}
\ON It is worth noting that the Jacobian \OFF  J has the physical meaning of being the local ratio of current to reference volume of a material volume element.

	\subsection{Change of surface}
For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration. Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation, the reference and current volumes are given respectively by:
\begin{align}
d\mathbf{V} &= d\mathbf{L} \cdot d\mathbf{A} \\
\mbox{and } \quad d\mathbf{v} &= d\mathbf{l} \cdot d\mathbf{a}.
\end{align}
Using \eqref{chap2:dvdV} which relates the reference and deformed volumes and recalling that $d\mathbf{l} = \mathbf{F} \, d\mathbf{L}$, we have:
\begin{equation}
J \, d\mathbf{L} \cdot d\mathbf{A} = (\mathbf{F} \, d\mathbf{L}) \cdot d\mathbf{a}.
\end{equation}
\ON Since \OFF this expression is valid for any vector $d\mathbf{L}$ and by using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$, \ON the sought relation between an element of area in the reference configuration $d\mathbf{A}$ and its corresponding area $d\mathbf{a}$ in the deformed configuration may be expressed as: \OFF
\begin{equation}
\label{chap2:dadA}
d\mathbf{a} = J \, \mathbf{F}^{-T} \, d\mathbf{A}.
\end{equation}
% 
%\begin{figure}[h]
%\begin{center}
%\includegraphics[width=13cm]{chapter2/changeSurface.pdf}
%\end{center}
%\caption{Transformation of a surface element under a deformation mapping. \TODO{CHANGE TO A BETTER FIGURE.}}
%\label{chap2:fig-changeSurface}
%\end{figure}

@@

	\subsection{Green-Lagrange strain tensor}
Using \eqref{chap2:dS} and \eqref{chap2:ds}, \ON the difference in the squared lengths between the reference and the current configuration may be evaluated as: \OFF 
\begin{align}
(ds)^2 - (dS)^2 &= d\mathbf{X} \cdot (\mathbf{C} \, d\mathbf{X}) - d\mathbf{X} \cdot d\mathbf{X} \notag \\
&= d\mathbf{X} \cdot  (\mathbf{C} - \mathbf{I}) \, d\mathbf{X}.
\end{align}
Let us define the \emph{Green-Lagrange strain tensor} $\mathbf{E}$ as:
\begin{equation}
\mathbf{E} = \frac{1}{2}	 (\mathbf{C} - \mathbf{I})
\end{equation}
so we can write:
\begin{equation}
(ds)^2 - (dS)^2 = 2 \, d\mathbf{X} \cdot \mathbf{E} \, d\mathbf{X}.
\end{equation}
By definition, the Green-Lagrange strain tensor is a symmetric second-order tensor. \ON Moreover, we observe that \OFF the change in squared lengths is zero if and only if \ON the Green-Lagrange strain tensor \OFF $\mathbf{E} = \mathbf{0}$. Using \eqref{chap2:gradient}, the Green strain tensor may be expanded as the following:
\begin{align}
\mathbf{E} &= \frac{1}{2} (\mathbf{F}^T \mathbf{F} - \mathbf{I}) \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u} + \mathbf{I})^T (\nabla_{0} \mathbf{u} + \mathbf{I}) - \mathbf{I} \right] \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right]. \label{chap2:GreenTensor}
\end{align}

@@

\section{Stress}
Stress is a measure of the intensity of the internal forces acting between particles of a deformable body across imaginary internal surfaces. These internal forces are produced between the particles in the body as a reaction to external forces applied on the body. The SI unit for stress is pascal (symbol $Pa$), which is equivalent to one newton (force) per square metre (unit area). As with strain, stress can be measured per unit of deformed area (section \ref{chap2:CauchyStress}) or undeformed area (sections \ref{chap2:PiolaStress1} and \ref{chap2:PiolaStress2}). In general, the stress is not uniformly distributed over the cross section of a material body, and consequently the stress at a point on a given area is different from the average stress over this entire area. Therefore, it is necessary to define the stress not over a given area but at a specific point in the body. \ON The stress at a point in a three-dimensional continuum is fully defined by \OFF nine quantities, three per plane, on three mutually perpendicular planes at the point. \ON The \emph{stress tensor} is defined as the second-order tensor which components are these nine quantities. \OFF

@@

\ON In the following we make use of a Cartesian coordinate system and we consider an infinitesimal tetrahedron as shown on \fig{chap2:fig-CauchyStress}. \OFF
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=6cm]{chapter2/CauchyStress.pdf}
\end{center}
\caption{Tetrahedral element in Cartesian coordinates.}
\label{chap2:fig-CauchyStress}
\end{figure}
%
We note $-\mathbf{t}_1$, $-\mathbf{t}_2$, $-\mathbf{t}_3$ and $\mathbf{t}$ the stress vectors in the outwards directions on the faces of the tetrahedron whose areas are respectively $\Delta a_1$, $\Delta a_2$, $\Delta a_3$ and $\Delta a$. For the mass $m$ inside the tetrahedron, we have by Newton's second law:
\begin{equation}
\label{chap2:Newton}
\sum \mathbf{F} = \mathbf{t} \Delta a - \mathbf{t}_1 \Delta a_1 - \mathbf{t}_2 \Delta a_2 - \mathbf{t}_3 \Delta a_3 + \Delta v \mathbf{f} = m \mathbf{a},
\end{equation}
where $\mathbf{f}$ is the force per unit volume acting on the body, $\Delta v$ the volume of the tetrahedron and $\mathbf{a}$ the acceleration.
The areas $\Delta a_1$, $\Delta a_2$ and $\Delta a_3$ being the projections of $\Delta a$, they are related to $\Delta a$ in the Cartesian coordinate system $(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ by:
\begin{equation}
\label{chap2:DeltaAi}
\Delta a_i = (\mathbf{n} \cdot \mathbf{e}_i) \Delta a, \quad i=1, 2, 3.
\end{equation}
Dividing \eqref{chap2:Newton} by $\Delta a$, using \eqref{chap2:DeltaAi} and noting that $\Delta v \slash \Delta a \rightarrow 0$ when the tetrahedron shrinks to a point yields:
\begin{align}
\mathbf{t} &= (\mathbf{n} \cdot \mathbf{e}_1) \mathbf{t}_1 + (\mathbf{n} \cdot \mathbf{e}_2) \mathbf{t}_2 + (\mathbf{n} \cdot \mathbf{e}_3) \mathbf{t}_3 \notag \\
&= \mathbf{n} \cdot (\mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3).
\end{align}

@@

The Cauchy stress tensor is used for stress analysis of material bodies experiencing small deformations. This stress is basically defined as force/(unit deformed area). The strain measure that is appropriate to use with the Cauchy stress tensor is \ON therefore \OFF the infinitesimal strain tensor (see section~\ref{chap2:infinitesimalStrainTensor}).  \ON Indeed, because the Cauchy stress tensor is defined by unit of deformed area, it is not appropriate for analysing materials undergoing large deformation where the area in the deformed configuration is often unknown. Consequently, we require alternative stress tensors based on the reference configuration. \OFF We will introduce two new stress measures: the first Piola-Kirchhoff stress tensor and the second Piola-Kirchhoff stress tensor.

	\subsection{First Piola-Kirchhoff stress tensor}\label{chap2:PiolaStress1}
We know that the Cauchy stress tensor represents the current force per unit of deformed area: $d\mathbf{f} = \mathbf{t}(\mathbf{n}) da = \sigma \cdot d\mathbf{a}$. \ON We state that we can find a stress vector $\mathbf{T}$ over the area element $d\mathbf{A}$ in the underformed configuration which \OFF results in the same total force.
\begin{equation}
\label{chap2:df}
d\mathbf{f} = \mathbf{t}(\mathbf{n}) da = \mathbf{T}(\mathbf{N}) dA.
\end{equation} 
The two stress vectors have the obviously the same direction but different magnitudes since they are not applied to the same area.

In a similar way we defined the Cauchy stress $\sigma$ with $\mathbf{t}(\mathbf{n}) = \sigma \cdot \mathbf{n}$, we introduce a stress tensor $\mathbf{P}$ such that $\mathbf{T}(\mathbf{n}) = \mathbf{P} \cdot \mathbf{N}$. \ON We call $ \mathbf{P} $ the \emph{first Piola-Kirchhoff stress tensor} and \OFF gives the current force per unit undeformed area.

@@

\section{Constitutive equations}
So far, all the relations that we have established are valid for every continuum. Indeed, \ON the derivations carried out so far made no mention of any material. \OFF Therefore, \ON we know that \OFF these relations are not sufficient to describe the response of a material to a given loading. \ON Moreover, because the stresses result from the deformation of the material, they may be expressed \OFF in terms of some measure of this deformation such as the strain. This relationship, which obviously depends on the type of material under consideration, is what we call a \emph{constitutive equation}. \ON Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimentals results. \OFF

>>>>

\chapter[Background in continuum mechanics for soft-tissue modelling][Background in continuum mechanics]{Background in continuum mechanics for soft-tissue modelling}
\label{chap2}
\begin{shortAbstract}
As seen in the previous chapter, realistic modelling of organ deformation is a challenging research field that opens the door to new clinical applications including: medical training and rehearsal systems, patient-specific planning of surgical procedures and per-operative guidance based on simulation. In all these cases the clinician needs fast updates of the deformation model to obtain a real-time display of the computed deformations. If for medical training devices the haptic feedback from touching organs merely needs to feel real, the accuracy of the information provided to the clinician in the cases of planning or per-operative guidance is crucial. Therefore, a substantial comprehension of the mechanics involved and a knowledge of the physical properties of anatomical structures are both mandatory in our quest to realistically model the deformation of organs. This chapter will start by introducing the main concepts of continuum mechanics that are fundamental to study the mechanical response of organs. It will then present mathematical models able to describe the different mechanical aspects of materials and briefly discuss the mechanical characterisation of tissues.
\end{shortAbstract}

@@

\section{Description of motion} 	\label{chap2:descriptionMotion}
Let us consider a body $\textbf{\textit{B}}$ of known geometry in a three-dimensional Euclidian space $\field{R}^3$. For a given geometry and loading, $\textbf{\textit{B}}$ will undergo a set of macroscopic changes which is called \emph{deformation}. The region of space occupied by the body at a given time $t$ is termed a \emph{configuration}. A change in the configuration of a continuum body results in a displacement. The displacement of a body has two components: a rigid-body displacement and a deformation. A rigid-body displacement consists of a simultaneous translation and rotation of the body without changing its shape or size. In constrast, a change in shape and\slash or size of the body from an initial configuration $\kappa_{0}$ to a new configuration $\kappa$ is called a deformation. This new configuration $\kappa$ may be designated by the \emph{current} or \emph{deformed configuration}. \OFF

@@

	\subsection{Lagrangian description}	
In the Lagrangian description, the position and physical properties of the particles are referred to a reference configuration $\kappa_{R}$, often chosen to be the undeformed configuration $\kappa_{0}$. Consequently, the current coordinates ($\mathbf{x} \in \kappa$) depend on the reference coordinates ($\mathbf{X} \in \kappa_{0}$): \OFF
\begin{equation}
\textbf{x} = \chi(\textbf{X}, t), \quad \chi(\textbf{X}, 0) = \textbf{X},
\end{equation}
and a typical variable $\phi$ defined over the body is expressed in terms of the coordinates $\mathbf{X}$ and time $t$: \OFF
\begin{equation}
\label{chap2:phi}
\phi = \phi(\textbf{X}, t).
\end{equation}
Let us consider a particule $X$ whose position in the reference configuration is $\mathbf{X}$. From \eqref{chap2:phi}, we note that the value of $\phi$ associated with this fixed particle $X$ changes over time. Hence, the Lagrangian description focuses its attention on the particles of the continuous body and it is usually used in solid mechanics.

	\subsection{Eulerian description}
In constrast, the Eulerian description is interested in changes at fixed locations. This time, the motion and $\phi$ are described with respect to the current position ($\mathbf{x} \in \kappa$): \OFF
\begin{equation}
\phi = \phi(\textbf{x}, t), \quad \textbf{X} = \textbf{X}(\textbf{x}, t).
\end{equation}
For a fixed value of $\mathbf{x} \in \kappa$, $\phi(\mathbf{x},t)$ gives the value of $\phi$ associated with the particle occupying the position $\textbf{x} \in \kappa$, which may very well be a different particle for each new time $ t $. Because a change in time $t$ implies that a different value $\phi$ is observed at the same spatial location $\mathbf{x} \in \kappa$, now probably occupied by a different particle, the Eulerian description is focused on a spatial position. This approach is convenient for the study of fluid flow where the kinematic property of greatest interest is the rate at which change is taking place rather than the shape of the body of fluid at a reference time \citep{Spencer80}. Because we are only interested in the study of solid bodies, the Lagrangian description will be used in the rest of the text.

	\subsection{Displacement field}
The displacement $ \mathbf{u} $ of a particle $X$ is called \emph{displacement vector} and its expression in the Lagrangian description is given by the following:
\begin{equation}
\mathbf{u}(\textbf{X}, t) = \mathbf{x}(\mathbf{X}, t) - \mathbf{X}.
\end{equation}
A \emph{displacement field} is a vector field of all displacement vectors for all particles in the body. Thus, the deformed configuration $\kappa$ may be obtained from the undeformed configuration $\kappa_{0}$ by merely adding the displacement field: $\kappa_{0}$: $\chi(\mathbf{X}, t) = \mathbf{X} + \mathbf{u}(\mathbf{X})$.

@@

	\subsection{Deformation gradient tensor}
The displacement field tells us how a particule displaces from the reference to the deformed configuration. More interestingly, we would like to know how a material line would deform (stretch and rotation) within this displacement field. Since the length of a material line $d\mathbf{X}$ can change when going to the deformed configuration as well as its orientation, we can say that $d\mathbf{X}$ deforms into $d\mathbf{x}$. We now seek the relation existing between $d\mathbf{x}$ in the deformed configuration and $d\mathbf{X}$ of the reference configuration. \OFF

@@

	\subsection{Change of volume}
At this point, we have seen how a deformation can affect a vector. We will now look into its effect on volumes. Our motivation comes from the need to write global equilibrium statements that involve integrals over volumes. Let us consider a volume in the reference configuration formed by three non-coplanar line elements $d\mathbf{X}^{(1)}$, $d\mathbf{X}^{(2)}$ and $d\mathbf{X}^{(3)}$ as can be seen on \fig{chap2:fig-changeVolume}.
% 
\begin{figure}[ht]
\begin{center}
\includegraphics[width=12cm]{chapter2/changeVolume.pdf}
\end{center}
\caption{Transformation of a volume element under a deformation mapping.}
\label{chap2:fig-changeVolume}
\end{figure}
%

The three vectors after deformation $d\mathbf{x}^{(1)}$, $d\mathbf{x}^{(2)}$ and $d\mathbf{x}^{(3)}$ can be obtained with:
\begin{equation}
\label{chap2:dxdX}
d\mathbf{x}^{(i)} = \mathbf{F} \, d\mathbf{X}^{(i)}, \quad i = 1, 2, 3.
\end{equation}
The volume of the parallelepiped that we will note $dV$  can be calculated using the triple product between the three vectors:
\begin{align}
dV &= d\mathbf{X}^{(1)} \cdot d\mathbf{X}^{(2)} \times d\mathbf{X}^{(3)} = (\mathbf{\hat{N}}_{1} \cdot \mathbf{\hat{N}}_{2} \times \mathbf{\hat{N}}_{3}) \, dX^{(1)} dX^{(2)} dX^{(3)} \notag \\
&= dX^{(1)} dX^{(2)} dX^{(3)}, \label{chap2:dV}
\end{align}
where $\mathbf{\hat{N}}_{i}$ is the unit vector along $d\mathbf{X}^{i}$. Similarly, we can compute the corresponding volume in the deformed configuration by:
\begin{align}
dv &= d\mathbf{x}^{(1)} \cdot d\mathbf{x}^{(2)} \times d\mathbf{x}^{(3)} \notag \\
&=  (\mathbf{F} \cdot \mathbf{\hat{N}}_{1}) \cdot (\mathbf{F} \cdot \mathbf{\hat{N}}_{2}) \times (\mathbf{F} \cdot \mathbf{\hat{N}}_{3}) \,  dX^{(1)} dX^{(2)} dX^{(3)} \quad \text{by \eqref{chap2:dxdX}} \notag \\
&= \det \mathbf{F} \, dX^{(1)} dX^{(2)} dX^{(3)}. \label{chap2:dv}
\end{align}
The determinant of $\mathbf{F}$ is called the \emph{Jacobian} and it is denoted by $J = \det \mathbf{F}$. Therefore, the relation between the deformed volume $ dv $ and the undeformed volume $ dV $ may be written as: \OFF
\begin{equation}
\label{chap2:dvdV}
dv = J dV.
\end{equation}
It is worth noting that the Jacobian  J has the physical meaning of being the local ratio of current to reference volume of a material volume element.

	\subsection{Change of surface}
For similar reasons, let us find the relation between an element of area in the reference configuration $d\mathbf{A}$ which becomes $d\mathbf{a}$ in the deformed configuration. Considering an arbitrary material line $d\mathbf{L}$ in the reference configuration and noting $d\mathbf{l}$ the same line after deformation, the reference and current volumes are given respectively by:
\begin{align}
d\mathbf{V} &= d\mathbf{L} \cdot d\mathbf{A} \\
\mbox{and } \quad d\mathbf{v} &= d\mathbf{l} \cdot d\mathbf{a}.
\end{align}
Using \eqref{chap2:dvdV} which relates the reference and deformed volumes and recalling that $d\mathbf{l} = \mathbf{F} \, d\mathbf{L}$, we have:
\begin{equation}
J \, d\mathbf{L} \cdot d\mathbf{A} = (\mathbf{F} \, d\mathbf{L}) \cdot d\mathbf{a}.
\end{equation}
Since this expression is valid for any vector $d\mathbf{L}$ and by using the property $\mathbf{a} \cdot \mathbf{b} = \mathbf{a} ^{T} \mathbf{b}$, the sought relation between an element of area in the reference configuration $d\mathbf{A}$ and its corresponding area $d\mathbf{a}$ in the deformed configuration may be expressed as: \OFF
\begin{equation}
\label{chap2:dadA}
d\mathbf{a} = J \, \mathbf{F}^{-T} \, d\mathbf{A}.
\end{equation}
% 
%\begin{figure}[h]
%\begin{center}
%\includegraphics[width=13cm]{chapter2/changeSurface.pdf}
%\end{center}
%\caption{Transformation of a surface element under a deformation mapping. \TODO{CHANGE TO A BETTER FIGURE.}}
%\label{chap2:fig-changeSurface}
%\end{figure}

@@

	\subsection{Green-Lagrange strain tensor}
Using \eqref{chap2:dS} and \eqref{chap2:ds}, the difference in the squared lengths between the reference and the current configuration may be evaluated as: 
\begin{align}
(ds)^2 - (dS)^2 &= d\mathbf{X} \cdot (\mathbf{C} \, d\mathbf{X}) - d\mathbf{X} \cdot d\mathbf{X} \notag \\
&= d\mathbf{X} \cdot  (\mathbf{C} - \mathbf{I}) \, d\mathbf{X}.
\end{align}
Let us define the \emph{Green-Lagrange strain tensor} $\mathbf{E}$ as:
\begin{equation}
\mathbf{E} = \frac{1}{2}	 (\mathbf{C} - \mathbf{I})
\end{equation}
so we can write:
\begin{equation}
(ds)^2 - (dS)^2 = 2 \, d\mathbf{X} \cdot \mathbf{E} \, d\mathbf{X}.
\end{equation}
By definition, the Green-Lagrange strain tensor is a symmetric second-order tensor. Moreover, we observe that the change in squared lengths is zero if and only if the Green-Lagrange strain tensor $\mathbf{E} = \mathbf{0}$. Using \eqref{chap2:gradient}, the Green strain tensor may be expanded as the following:
\begin{align}
\mathbf{E} &= \frac{1}{2} (\mathbf{F}^T \mathbf{F} - \mathbf{I}) \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u} + \mathbf{I})^T (\nabla_{0} \mathbf{u} + \mathbf{I}) - \mathbf{I} \right] \notag \\
&= \frac{1}{2} \left[ (\nabla_{0} \mathbf{u})^T  + \nabla_{0} \mathbf{u}+ (\nabla_{0} \mathbf{u})^T(\nabla_{0} \mathbf{u}) \right]. \label{chap2:GreenTensor}
\end{align}

@@

\section{Stress}
Stress is a measure of the intensity of the internal forces acting between particles of a deformable body across imaginary internal surfaces. These internal forces are produced between the particles in the body as a reaction to external forces applied on the body. The SI unit for stress is pascal (symbol $Pa$), which is equivalent to one newton (force) per square metre (unit area). As with strain, stress can be measured per unit of deformed area (section \ref{chap2:CauchyStress}) or undeformed area (sections \ref{chap2:PiolaStress1} and \ref{chap2:PiolaStress2}). In general, the stress is not uniformly distributed over the cross section of a material body, and consequently the stress at a point on a given area is different from the average stress over this entire area. Therefore, it is necessary to define the stress not over a given area but at a specific point in the body. The stress at a point in a three-dimensional continuum is fully defined by nine quantities, three per plane, on three mutually perpendicular planes at the point. The \emph{stress tensor} is defined as the second-order tensor which components are these nine quantities.

@@

In the following we make use of a Cartesian coordinate system and we consider an infinitesimal tetrahedron as shown on \fig{chap2:fig-CauchyStress}. \OFF
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=6cm]{chapter2/CauchyStress.pdf}
\end{center}
\caption{Tetrahedral element in Cartesian coordinates.}
\label{chap2:fig-CauchyStress}
\end{figure}
%
We note $-\mathbf{t}_1$, $-\mathbf{t}_2$, $-\mathbf{t}_3$ and $\mathbf{t}$ the stress vectors in the outwards directions on the faces of the tetrahedron whose areas are respectively $\Delta a_1$, $\Delta a_2$, $\Delta a_3$ and $\Delta a$. For the mass $m$ inside the tetrahedron, we have by Newton's second law:
\begin{equation}
\label{chap2:Newton}
\sum \mathbf{F} = \mathbf{t} \Delta a - \mathbf{t}_1 \Delta a_1 - \mathbf{t}_2 \Delta a_2 - \mathbf{t}_3 \Delta a_3 + \Delta v \mathbf{f} = m \mathbf{a},
\end{equation}
where $\mathbf{f}$ is the force per unit volume acting on the body, $\Delta v$ the volume of the tetrahedron and $\mathbf{a}$ the acceleration.
The areas $\Delta a_1$, $\Delta a_2$ and $\Delta a_3$ being the projections of $\Delta a$, they are related to $\Delta a$ in the Cartesian coordinate system $(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ by:
\begin{equation}
\label{chap2:DeltaAi}
\Delta a_i = (\mathbf{n} \cdot \mathbf{e}_i) \Delta a, \quad i=1, 2, 3.
\end{equation}
Dividing \eqref{chap2:Newton} by $\Delta a$, using \eqref{chap2:DeltaAi} and noting that $\Delta v \slash \Delta a \rightarrow 0$ when the tetrahedron shrinks to a point yields:
\begin{align}
\mathbf{t} &= (\mathbf{n} \cdot \mathbf{e}_1) \mathbf{t}_1 + (\mathbf{n} \cdot \mathbf{e}_2) \mathbf{t}_2 + (\mathbf{n} \cdot \mathbf{e}_3) \mathbf{t}_3 \notag \\
&= \mathbf{n} \cdot (\mathbf{e}_1 \mathbf{t}_1 + \mathbf{e}_2 \mathbf{t}_2 + \mathbf{e}_3 \mathbf{t}_3).
\end{align}

@@

The Cauchy stress tensor is used for stress analysis of material bodies experiencing small deformations. This stress is basically defined as force/(unit deformed area). The strain measure that is appropriate to use with the Cauchy stress tensor is therefore the infinitesimal strain tensor (see section~\ref{chap2:infinitesimalStrainTensor}).  Indeed, because the Cauchy stress tensor is defined by unit of deformed area, it is not appropriate for analysing materials undergoing large deformation where the area in the deformed configuration is often unknown. Consequently, we require alternative stress tensors based on the reference configuration. We will introduce two new stress measures: the first Piola-Kirchhoff stress tensor and the second Piola-Kirchhoff stress tensor.

	\subsection{First Piola-Kirchhoff stress tensor}\label{chap2:PiolaStress1}
We know that the Cauchy stress tensor represents the current force per unit of deformed area: $d\mathbf{f} = \mathbf{t}(\mathbf{n}) da = \sigma \cdot d\mathbf{a}$. We state that we can find a stress vector $\mathbf{T}$ over the area element $d\mathbf{A}$ in the underformed configuration which results in the same total force.
\begin{equation}
\label{chap2:df}
d\mathbf{f} = \mathbf{t}(\mathbf{n}) da = \mathbf{T}(\mathbf{N}) dA.
\end{equation} 
The two stress vectors have the obviously the same direction but different magnitudes since they are not applied to the same area.

In a similar way we defined the Cauchy stress $\sigma$ with $\mathbf{t}(\mathbf{n}) = \sigma \cdot \mathbf{n}$, we introduce a stress tensor $\mathbf{P}$ such that $\mathbf{T}(\mathbf{n}) = \mathbf{P} \cdot \mathbf{N}$. We call $ \mathbf{P} $ the \emph{first Piola-Kirchhoff stress tensor} and gives the current force per unit undeformed area.

@@

\section{Constitutive equations}
So far, all the relations that we have established are valid for every continuum. Indeed, the derivations carried out so far made no mention of any material. Therefore, we know that these relations are not sufficient to describe the response of a material to a given loading. Moreover, because the stresses result from the deformation of the material, they may be expressed in terms of some measure of this deformation such as the strain. This relationship, which obviously depends on the type of material under consideration, is what we call a \emph{constitutive equation}. Constitutives equations are an attempt to mathematically characterise the behaviour of materials observed from experimentals results. \OFF

Naturally, the relationship between strain and stress is conditioned by the constitution of the material of interest. Because the mechanics involved in materials can be quite complex (tissue mechanics in particular), many simplifying assumptions are required to derive a constitutive equation. In practice, a given constitutive equation can only cover a subset of the actual mechanical behaviour and could not possibly model all aspects of tissue behaviour under any type of loading accurately. For instance, some tissues behave very differently under small loads as compared to large loads. The behaviour may also depends on loading rates and damage may even occur under large loads. Moreover, not only constitutive equations need to deal with the complexity of material mechanics but they must remain simple enough to measure all the included material constants experimentally. This section will now introduce the most common constitutive models relevant to soft-tissue modelling.
